---
title: "Functions"

output: pdf_document
---

## 6.1 Function components

All R functions have three parts:

*  the `body()`, the code inside the function.
*  the `formals()`, the list of arguments which controls how you can call the function.
*  `the environment()`, the “map” of the location of the function’s variables.

These three important components will show up when print a function in R.  If the environment isn’t displayed, it means that the function was created in the global environment.

```{r}
f <- function(x) x^2
f

formals(f)

body(f)

environment(f)

```

The assignment forms of body(), formals(), and environment() can also
be used to modify functions.

Like all objects in R, functions can also possess any number of additional
`attributes()`. One attribute used by base R is “srcref”, short for source
reference, which points to the source code used to create the function.
Unlike `body()`, this contains code comments and other formatting. 

### 6.1.1 Primitive functions

There is one exception to the rule that functions have three components.
Primitive functions, like `sum()`, call `C` code directly with `.Primitive()` and contain no `R` code. Therefore their `formals()`, `body()`, and
`environment()` are all NULL:

```{r}
sum

formals(sum)

body(sum)

environment(sum)

```

Primitive functions are only found in the base package, and since they
operate at a low level, they can be more efficient (primitive replacement
functions don’t have to make copies), and can have different rules for
argument matching (e.g., switch and call). 

### 6.1.2 Exercises

1.  What function allows you to tell if an object is a function? What function allows you to tell if a function is a primitive function?

`is.function` and `is.primitive`

2.  This code makes a list of all functions in the base package.

    objs <- mget(ls("package:base"), inherits = TRUE)  
    funs <- Filter(is.function, objs)
    Use it to answer the following questions:
      a.  Which base function has the most arguments?
      b.  How many base functions have no arguments? What’s special about those functions?
      c.  How could you adapt the code to find all primitive functions?
      
```{r}
objs <- mget(ls("package:base"), inherits = TRUE)  
funs <- Filter(is.function, objs)
# a. :
x <- numeric(length(funs))
for (i in 1: length(funs)){
    x[i] <- length(formals(funs[[i]]))
    }
max(x)
funs[which(x == max(x))]

# b. :
length(which(x == 0)) #224
# These are primitive functions that does not have formals

# c. :
findPrimitive <- function(){
    
    # find all functions in base package
    objs <- mget(ls("package:base"), inherits = TRUE)  
    funs <- Filter(is.function, objs)
    
    # find length of all functions
    x <- numeric(length(funs))
        for (i in 1: length(funs)){
        x[i] <- length(formals(funs[[i]]))
    }
    
    # find the number of primitive functions
    n <- length(which(x == 0))
    
    # number of all primitive functions
    num <- which(x == 0)
    results <- list(0)
    results <- funs[num]
    return(results)
}

```

3.  What are the three important components of a function?

body, formals and environment

4.  When does printing a function not show what environment it
was created in?

It means that the function was created in the global environment.

## 6.2 Lexical scoping

Scoping is the set of rules that govern how R looks up the value of a symbol.

R has two types of scoping: lexical scoping, implemented automatically at the language level, and dynamic scoping, used in select functions to save typing during interactive analysis.

Lexical scoping looks up symbol values based on how functions were nested when they were created, not how they are nested when they are called.

There are four basic principles behind R’s implementation of lexical scoping:

*  name masking  
*  functions vs. variables  
*  a fresh start  
*  dynamic lookup  

### 6.2.1 Name masking

The following example illustrates the most basic principle of lexical scoping

```{r}
f <- function(){
    x <- 1
    y <- 2
    c(x, y)
}
f()
rm(f)

# If a name isn’t defined inside a function, R will look one level up
x <- 2
g <- function(){
    y <- 1
    c(x, y)
}
g()
rm(x, g)
```

The same rules apply if a function is defined inside another function:  
look inside the current function, then where that function was defined, and so on, all the way up to the global environment, and then on to other loaded packages.

```{r}
x <- 1
h <- function(){
    y <- 2
    i <- function(){
        z <- 3
        c(x, y, z)
    }
    i()
}
h()
rm(x, h)
```

The same rules apply to closures, functions created by other functions. The following function, `j()`, returns a function. 

```{r}
j <- function(x){
    y <- 2
    function(){
        c(x, y)
    }
}
k <- j(1)
k()
rm(j, k)
```

### 6.2.2 Functions vs. variables

The same principles apply regardless of the type of associated value — finding functions works exactly the same way as finding variables:

```{r}
l <- function(x) x + 1
m <- function(){
    l <- function(x) x * 2
    l(10)
}
m()
rm(l, m)
```

### 6.2.3 A fresh start

What happens to the values in between invocations of a function? What will happen the first time you run this function? What will happen the second time?

```{r}
j <- function(){
    if(!exists("a")) {
        a <- 1
    } else{
        a <- a + 1
    }
    print(a)
}
j()
rm(j)
# This is because every time a function is called, a new environment is created to host execution.
```

### 6.2.4 Dynamic lookup

Lexical scoping determines where to look for values, not when to look for them. R looks for values when the function is run, not when it’s created.

```{r}
f <- function() x
x <- 15
f()

x <- 20
f()
#  this means the function is no longer self-contained
```

One way to detect this problem is the `findGlobals()` function from codetools. This function lists all the external dependencies of a function:

```{r}
f <- function() x + 1
codetools::findGlobals(f)
```

Another way to try and solve the problem would be to manually change the environment of the function to the `emptyenv()`, an environment which contains absolutely nothing:

```{r}
environment(f) <- emptyenv()
f()
```

### 6.2.5 Exercises

1.  What does the following code return? Why? What does each of the three c’s mean?

c <- 10  
here the c is a scalar  

c(c = c)  
The first c, c(), is a generic function that combines its arguments to form a vector, the second c is the name of the argument, the third c represent the variable c which equals to 10.

2.  What are the four principles that govern how R looks for values?

The four priciples are:    

* Name masking
* Function vs. variables
* a fresh start
* dynamic lookup

3. What does the following function return? Make a prediction before running the code yourself.

```{r}
f <- function(x) {
    f <- function(x) {
        f <- function(x) {
            x ^ 2
            }
        f(x) + 1
    }
    f(x) * 2
}
f(10)
```

### 6.3 Every operation is a function call

The previous example of redefining ( works because every operation in R is a function calle. This includes infix operators like `+`, control flow operators like `for`, `if`, and `while`, subsetting operators like `[]` and `$`, and even the curly brace `{`. This means that each pair of statements in the following example is exactly equivalent.

```{r}
x <- 10; y <- 5
x + y

`+`(x, y)

for (i in 1:2) print(i)
`for`(i, 1:2, print(i))

if (i == 1) print("yes!") else print("no")
`if` (i == 1, print("yes!"), print("no."))

x[3]
`[`(x, 3)

# { print(1); print(2); print(3) }
{print(1); print (2); print(3)}
`{`(print(1), print(2), print(3))
```

It’s more often useful to treat special functions as ordinary functions.

```{r}
add <- function(x, y) x + y
sapply(1:10, add, 3)

# can also get the same effect using the built-in + function
sapply(1:10, `+`, 3)
sapply(1:10, "+", 3)

# combine lapply() or sapply() with subsetting:
x <- list(1:3, 4:9, 10:12)
sapply(x, "[", 2)

# equivalent to
sapply(x, function(x) x[2])
```

## 6.4 Function arguments

The formal arguments are a property of the function, whereas the actual or calling arguments can vary each time you call the function.

### 6.4.1 Calling functions

Can specify arguments by position, by complete name, or by partial name when calling a function. Arguments are matched first by exact name (perfect matching), then by prefix matching, and finally by position.

```{r}
f <- function(abcdef, bcde1, bcde2) {
    list(a = abcdef, b1 = bcde1, b2 = bcde2)
}
str(f(1, 2, 3))

str(f(2, 3, abcdef = 1))

# Can abbreviate long argument names:
str(f(2, 3, a = 1))

# But this doesn't work because abbreviation is ambiguous
str(f(1, 3, b = 1))
```

Avoid using positional matching for less commonly used argumentss, and only use readable abbreviations with partial matching. Named arguments should always come after unnamed arguments. 

These are good calls:

mean(1:10)  
mean(1:10, trim = 0.05)  

These are confusing:

mean(1:10, n = T)  
mean(1:10, , FALSE)  
mean(1:10, 0.05)  
mean(, TRUE, x = c(1:10, NA))  

### 6.4.2 Calling a function given a list of arguments

Suppose a list of function arguments:

```{r}
args <- list(1:10, na.rm = TRUE)

# How to use functions mean() (do.call)
do.call(mean, list(1:10, na.rm = TRUE))

# Equivalent to
mean(1:10, na.rm = TRUE)
```

### 6.4.3 Default and missing arguments

Function arguments in R can have default values.

```{r}
f <- function(a = 1, b = 2){
    c(a, b)
}
f()

# the default value can be defined in terms of other arguments:
g <- function(a = 1, b = a*2) {
    c(a, b)
}
g()
g(10)

# Default arguments can even be defined in terms of variables created within the function. 
h <- function(a = 1, b = d) {
    d <- (a + 1)^2
    c(a, b)
}
h()
h(10)

# can determine if an argument was supplied or not with the missing() function.
i <- function(a, b) {
    c(missing(a), missing(b))
}
i()
i(a = 1)
i(b = 2)
i(1, 2)
```

### 6.4.4 Lazy evaluation

By default, R function arguments are only evaluated if they’re actually used:

```{r}
f <- function(x) {
    10
}
f(stop("This is an error"))

# use force() to ensure that an argument is evaluated
f <- function(x) {
    force(x)
    10
} 
f(stop("This is an error"))

# This is important when creating closures with lapply() or a loop:
add <- function(x) {
    function(y) x + y
}
adders <- lapply(1:10, add)
adders[[1]](10)
adders[[10]](10)

add <- function(x) {
    force(x)
    function(y) x + y
}
adders2 <- lapply(1:10, add)
adders2[[1]](10)
adders2[[10]](10)

# This code is exactly equivalent to
add <- function(x) {
    x
    function(y) x + y
}
```

Default arguments are evaluated inside the function.

```{r}
f <- function(x = ls()) {
    a <- 1
    x
}
# ls() evaluated inside f:
f()

# ls() evaluated in global environment:
ls()
```

an unevaluated argument is called a promise, or  a thunk. A promise is made up of two parts:

*  The expression which gives rise to the delayed computation. (It can be accessed with `substitute()`. )
*   The environment where the expression was created and where it should be evaluated

The first time a promise is accessed the expression is evaluated in the environment where it was created. This value is cached, so that subsequent access to the evaluated promise does not recompute the value (but the
original expression is still associated with the value, so `substitute()` can
continue to access it).


Laziness is useful in if statements — the second statement below will
be evaluated only if the first is true. If it wasn’t, the statement would
return an error because NULL > 0 is a logical vector of length 0 and not
a valid input to if.

```{r}
x <- NULL
if (!is.null(x) && x > 0) {
    
}

# implement &&
`&&` <- function(x, y) {
    if(!x) return(FALSE)
    if(!y) return(FALSE)
    
    TRUE
}
a <- NULL
!is.null(a) && a > 0
```

This function would not work without lazy evaluation because both x and y would always be evaluated, testing a > 0 even when a was NULL.

### 6.4.5 ...

This argument will match any arguments not otherwise matched, and can be easily passed on to other functions. One relatively sophisticated user of ... is the base `plot()` function. `plot()` is a generic method with arguments x, y and ... .  ... accepts “other graphical parameters”, which are listed in the help for par().

use `list(...)` to capture ... in a form that is easier to work with.

```{r}
f <- function(...) {
    names(list(...))
}

f(a = 1, b = 2)
```

When using ..., any misspelled arguments will not raise an error, and any arguments after ... must be fully named.

### 6.4.6 Exercises

1.  Clarify the following list of odd function calls:

```{r}
x <- sample(replace = TRUE, 20, x = c(1:10, NA))
# randomly sample 20 elements from numbers 1 to 10 and NA, with replacement.

y <- runif(min = 0, max = 1, 20)
# simulate 20 numbers between 0 and 1

cor(m = "k", y = y, u = "p", x = x)
# using pearson and pairwise.complete.obs methods to calculate
```

2.  What does this function return? Why? Which principle does it illustrate?

```{r}
f1 <- function(x = {y <- 1; 2}, y = 0) {
x + y
}
f1()
```

3.  What does this function return? Why? Which principle does it illustrate?

```{r}
f2 <- function(x = z) {
z <- 100
x
}
f2()
```

## 6.5 Special calls

### 6.5.1 Infix functions

Most functions in R are “prefix” operators: the name of the function comes before the arguments. Infix functions are functions that function name comes in between its arguments, like + or -. All user created infix functions must start and end with % and R comes with the following infix functions predefined: %%, %*%, %/%, %in%, %o%, %x%. (The complete list of built-in infix operators that don’t need % is: ::, :::, $, @, ˆ, *, /, +, -, >, >=, <, <=, ==, !=, !, &, &&, |, ||, ~, <-, <<-)

Example, create a new operator that pastes together strings:

```{r}
`%+%` <- function (a, b) paste(a, b, sep = "")
"new" %+% " string"
```

When creating the function, put the function name in backticks because it’s a special name.

## 6.6 Return values

The last expression evaluated in a function becomes the return value, the result of invoking the function.

```{r}
f <- function(x){
    if (x < 10){
        0
    } else {
        10
    }
}
f(5)
f(15)
```

Functions can return only a single object but can return a list containing any number of objects. R protects you from one type of side effect: most R objects have copy-onmodify semantics. So modifying a function argument does not change the original value:

```{r}
f <- function(x) {
    x$a <- 2
    x
}
x <- list(a = 1)
f(x)
x$a
```

### 6.6.1 On exit

Functions can set up other triggers to occur when the function is finished using `on.exit()`.

```{r}
in_dir <- function(dir, code) {
    old <- setwd(dir)
    on.exit(setwd(old))
    force(code)
}
getwd()
in_dir("~", getwd())
```
