---
title: "Data Structures"
output: pdf_document
---

## Introduction

R's base data structures can be organized by their dimensionality (1d, 2d, or nd) and whether they're homogeneous (all contents must be of the same type) or heterogeneous (the contents can be of different types). This gives rise the five data types most often used in data analysis:

|    | Homogeneous     |  Heterogeneous  |
|----|:----------------|:----------------|
| 1d | Atomic vector   | List            |
| 2d | Matrix          | Data frame      |
| nd | Array

Almost all other objects are built upon these foundations. Note that R has no 0-dimentional, or scalar types. Individual numbers or strings are actually vectors of length one. Given an object, the best way to understand what data structures it's composed of is to use `str()`.

## 1.1 Vectors
The basic data structure in R is the vector. There are two kinds of vectors: atomic vectors and lists. They have three common properties:   
* Type, typeof(), what it is.
* Length, length(), how many elements it contains.
* Attributes, attributes(), additional arbitrary metadata.

They differ in the types of their elements: all elements of an atomic vector must be the same type, whereas the elements of a list can have different types. Use `is.vector() || is.list()` to test if an object actually a vector.

## 1.1.1 Atomic Vector
There are four common types of atomic vectors: logical, integer, double(often called numeric), and character.

Create atomic vectors with `c()`:
```{r}
dbl_var <- c(1, 2.5, 4.5)
int_var <- c(1L, 6L, 10L)# get an integer vector with L suffix
log_var <- c(TRUE, FALSE, T, F) # use TRUE, FALSE(or T, F) to create logical vector
chr_var <- c("these are", "some strings")

#Atomic vectors are always flat
c(1, c(2, c(3, 4)))
```

### 1.1.1.1 Types and tests
We can determine a vector's type with `typeof()`, or check if it is a specific type with an "is" function: `is.character()`, `is.double()`, `is.integer()`, `is.logical()`, or `is.atomic()`.

```{r}
typeof(int_var)
is.integer(int_var)
is.atomic(int_var)

typeof(dbl_var)
is.double(dbl_var)
is.atomic(dbl_var)
```

`is.numeric` is a general test and return *TRUE* for both integer and double vectors.

### 1.1.1.2 Coercion
If we combine two types of vectors they will be *coerced* to the most flexible type. Types from least to most flexible are: logical, integer, double and character. When a logical vector is coerced to an integer or double, *TRUE* becomes 1 and *FALSE* becomes 0. For example, combining a character and an integer yields a character:
```{r}
str(c("a", 1))
x <- c(FALSE, FALSE, TRUE)
as.numeric(x)

# Total number of TRUEs
sum(x)

# Proportion that are TRUE
mean(x)
```
## 1.1.2 Lists
The elements in a list can be any type, including lists, and we can construct lists using `list()`

```{r}
x <- list(1:3, "a", c(TRUE, FALSE, TRUE), c(2.3, 5.9))
str(x)
```
Lists are sometimes called *recursive* vectors, because a list can contain other lists.
```{r}
x <- list(list(list(list())))
str(x)
is.recursive(x)
# if given a combination of atomic vectors and lists, c() will #coerce the vector to list before combining them.
x <- list(list(1, 2), c(3, 4))
y <- c(list(1, 2), c(3, 4))
str(x)
str(y)
```
`typeof()` a list is list, test for a list with `is.list()`, coerce to a list with `as.list()`, turn a list into an atomic vector with `unlist()` using the same coersion rules as `c()`.

Lists are used to build up many of the more complicated data structures in R. For example, both data frames and linear models objects (as produced by `lm()`) are lists:
```{r}
is.list(mtcars)

mod <- lm(mpg ~ wt, data = mtcars)
is.list(mod)
```

## 1.1.3 Exercises

1. What are the six types of atomic vector? How does a list differ from an atomic vector?

    Answer: The six types of atomic vector are:
    
     * logical (may contain TRUE, FALSE, NA)
     * integer
     * double (often called numeric)
     * complex
     * character
     * raw
    
2. What makes `is.vector()` and `is.numeric()` fundamentally different to `is.list()` and `is.character()`?

    Answer: 
    
     (1) `is.vector` does not test if an object is a vector, it returns *TRUE* only if a vector has no other attributes than names. we can use `is.atomic() || is.list()` to test if an object is a vector.
     (2) `is.numeric()` returns *TRUE* for both integer and double vectors.
     (3) `is.list()` and `is.character()` are for testing if an object is a list or character vector.
 
3. Predict the output of the following uses of c() according to vector coercion rules.
 
 c(1, FALSE)  
 c("a", 1)  
 c(list(1), "a")  
 c(TRUE, 1L)  
 
    Answer: The coercion rules are two different types elements will be *coerced* to the most flexible type. Types from least to most flexible are: logical, integer, double and character. So,
    
     * c(1, FALSE) -> double  
     * c("a", 1) -> character  
     * c(list(1), "a") -> list  
     * c(TRUE, 1L) -> integer  

4. Why do you need to use `unlist()` to convert a list to an atomic vector? Why doesn't `as.vector()` work?

    Answer: a list actually is a vector, so if we use `as.vector()` on a list we get the list itself, so we need to use `unlist()` to convert a list to an atomic vector.

5. Why is 1 == "1" true? Why is -1 < FALSE true? Why is "one" < 2 false?

    Answer: when doing the comparison, 1 automatically coerced to "1", so "1" == "1". Similarly, FALSE coerced to 0 and compared to -1. Finally, 2 coerced to character before compare to "one", so now we compare two characters like sorting and we actually compare "2" and "o". Charcters are translated to UTF-8 before comparison, in UTF-8 coding "2" is 50 and "o" is "111", so "one" is greater than 2.

6. Why is the default missing value, NA, a logical vector? What's special about logical vectors? 

    Missing values are specified with NA, which is a logical vector of length
    1. NA will always be coerced to the correct type if used inside c(), or
    we can create NAs of a specific type with `NA_real_` (a double vector), `NA_integer_` and `NA_character_`.

## 1.2 Attributes

All objects can have arbitrary additional attributes, used to store metadata about the object. Attributes can be thought of a list of unique names. Attributes can be accessed individually with `attr()` or all at once as a list with `attributes()`.

```{r}
y <- 1:10
attr(y, "my_attribute") <- "This is a vector"
attr(y, "my_attribute")
str(attributes(y))
```

The `structure()` function returns a new object with modified attributes: 
```{r}
structure(1:10, my_attribute = "This is a vector")
```

By default, most attributes are lost when modifying a vector:
```{r}
attributes(y[1])
attributes(sum(y))
```

The only attributes not lost are three most important:

 * Names, a character vector giving each element a name.
 * Dimensions, used to turn vectors into matrices and arrays.
 * Class, used to implement the S3 object system.
 
Each of these attributes has a specific accessor function to get and set values: `names(x)`, `class(x)`, and `dim(x)`. 

You can name a vector in three ways:

  * When creating it: x <- c(a =1, b=2, c=3)
  * By modifying an existing vector in place: x <- 1:3; names(x) <- c("a", "b", "c")
  * By creating a modified copy of a vector: x <- setNames(1:3, c("a", "b", "c"))

We can create a new vector without names using `unname(x)`, or remove names in place with `names(x) <- NULL`.

### 1.2.1 Factors

One important use of attributes is to define factors. A factor is a vector that can only contain predefined values, and is used to store categorical data. Factors are built on top of integer vectors using two attributes: the *class()*, "factor", which makes them behave differently from regular integer vectors, and the *levels()*, which defines the set of allowed values.
```{r}
x <- factor(c("a", "b", "b", "a"))
x
class(x)
levels(x)

# cannot use values that are not in levels
x[2] <- "c"
x
# cannot combine vectors
c(factor("a"), factor("b"))
```

Factors are useful when you know the possible values a variable may take, and using a factor instead of a character vector makes it obvious when some groups contain no observations:
```{r}
sex_char <- c("m", "m", "m")
sex_factor <- factor(sex_char, levels = c("m", "f"))

table(sex_char)
table(sex_factor)
```

Sometimes when a data frame is read directly from a file, a column you’d
thought would produce a numeric vector instead produces a factor. This
is caused by a non-numeric value in the column, often a missing value
encoded in a special way like . or -. To remedy the situation, coerce the
vector from a factor to a character vector, and then from a character to
a double vector.
```{r}
z <- read.csv(text = "value\n12\n1\n.\n9")
typeof(z$value)
as.double(z$value)
class(z$value)
as.double(as.character(z$value))
z <- read.csv(text = "value\n12\n1\n.\n9", na.strings=".")
typeof(z$value)
class(z$value)
```

Most data loading functions in R automatically convert character vectors to factors. we can then use the argument `stringsAsFactors = FALSE` to suppress this behaviour, and then manually convert character vectors to factors using your knowledge of the data.

While factors look (and often behave) like character vectors, they are
actually integers. Be careful when treating them like strings. Some
string methods (like `gsub()` and `grepl()`) will coerce factors to strings, while others (like `nchar()`) will throw an error, and still others (like `c())` will use the underlying integer values. For this reason, it’s usually best to explicitly convert factors to character vectors if you need string-like behaviour.

### 1.2.2 Exercises

1. An early draft used this code to illustrate structure():
 
    ```
    structure(1:5, comment = "my attribute")
    #> [1] 1 2 3 4 5
    ```
    But when you print that object you don’t see the comment attribute. Why? Is the attribute missing, or is there something else special about it? 
    
    Answer: Unlike other attributes, the comment is not printed by `print` function. Use function `comment` to query or set a `comment` attribute.
    ```{r}
    comment(structure(1:5, comment = "my attribute"))
    ```

2. What happens to a factor when you modify its levels?
 
    ```
    f1 <- factor(letters)
    levels(f1) <- rev(levels(f1))
    ```
    
    The values in the factor variable will be reversed.

3. What does this code do? How do f2 and f3 differ from f1?
 
    ```
    f2 <- rev(factor(letters))
    f3 <- factor(letters, levels = rev(letters))
    ```
    f2 has the same values as the f1 but reversed levels, f3 has the levels as the f1 but reversed values.

## 1.3 Matrices and Arrays
Adding a dim() attribute to an atomic vector allows it to behave like a multi-dimensional array. A special case of the array is the matrix, which has two dimensions.

Matrices and arrays are created with matrix() and array(), or by using
the assignment form of dim():
```{r}
# Two scalar arguments to specify rows and columns
a <- matrix(1:6, ncol = 3, nrow = 2)
# One vector argument to describe all dimensions
b <- array(1:12, c(2, 3, 2))
a;b

# You can also modify an object in place by setting dim()
c <- 1:6
dim(c) <- c(3, 2)
c
dim(c) <- c(2, 3)
c
```

`length()` and `names()` have high-dimensional generalisations:

 * `length()` generalises to `nrow()` and `ncol()` for matrices, and `dim()` for arrays.
 * `names()` generalises to `rownames()` and `colnames()` for matrices, and
`dimnames()`, a list of character vectors, for arrays.

```{r}
length(a)
nrow(a)
ncol(a)
rownames(a) <- c("A", "B")
colnames(a) <- c("a", "b", "c")
a

length(b)
dim(b)
dimnames(b) <- list(c("one", "two"), c("a", "b", "c"), c("A", "B"))
b
```

`c()` generalises to `cbind()` and `rbind()` for matrices, and to `abind()` (provided by the abind package) for arrays. You can transpose a matrix with
`t()`; the generalised equivalent for arrays is `aperm()`.

Test if an object is a matrix or array using `is.matrix()` and `is.array()`, or by looking at the length of the `dim()`. `as.matrix()` and `as.array()` turn an existing vector into a matrix or array.

Can have matrices with a single row or single column, or arrays with a single dimension. They will behave differently from vectors, use str() to reveal the differences.

```{r}
str(1:3)            # 1d vector
str(matrix(1:3, ncol = 1)) # column vector
str(matrix(1:3, nrow = 1)) # row vector
str(array(1:3, 3)) # "array" vector
```

The dimension attribute can also be set on lists to make list-matrices or listarrays:
```
l <- list(1:3, "a", TRUE, 1.0)
dim(l) <- c(2, 2)
l
```

### 1.3.1 Exercises

  1. What does dim() return when applied to a vector?
 
    ```
    It will return "NULL". The dim() function retrieves or sets the dimension of an R object, 
    for example a matrix, array or data frame. For an array (and hence in particular, for a matrix) 
    dim retrieves the dim attribute of the object. It is NULL or a vector of mode integer.
    ```
  2. If is.matrix(x) is TRUE, what will is.array(x) return? 
 
    ```
    is.array(x) will also return TRUE. Matrix is a special case of the array.
    ```
 
  3. How would you describe the following three objects? What makes them different to 1:5?
 
    ```{r}
    x1 <- array(1:5, c(1, 1, 5))
    x2 <- array(1:5, c(1, 5, 1))
    x3 <- array(1:5, c(5, 1, 1))
    ## These are three dimensional arrays.
    ```

## 1.4 Data frames
A data frame is a list of equal-length vectors. This makes it a 2-dimensional structure, so it shares properties of both the matrix and the list. This means that a data frame has 'names()', 'colnames()', and 'rownames()'. The 'length()' of a data frame is the length of the underlying list and so is the same as 'ncol()'; 'nrow()' gives the number of rows.

### 1.4.1 Creation

create a data frame using data.frame() takes named vectors as input:
```{r}
df <- data.frame(x = 1:3, y = c("a", "b", "c"))
str(df)
```

Beware data.frame()’s default behaviour which turns strings into factors. Use stringAsFactors = FALSE to suppress this behaviour:
```{r}
df <- data.frame(
        x = 1:3,
        y = c("a", "b", "c"),
        stringsAsFactors = FALSE)
str(df)
```

### 1.4.2 Testing and coercion

Because a data.frame is an S3 class, its type reflects the underlying vector used to build it: the list. To check if an object is a data frame, use `class()` or test explicitly with `is.data.frame()`:
```{r}
typeof(df)
class(df)
is.data.frame(df)
```

Coerce an object to a data frame with `as.data.frame()`:

* A vector will create a one-column data frame.
* A list will create one column for each element; it’s an error if they’re
not all the same length.
* A matrix will create a data frame with the same number of columns
and rows.

### 1.4.3 Combining data frames

Combine data frames using `cbind()` and `rbind()`:
```{r}
cbind(df, data.frame(z = 3:1))
rbind(df, data.frame(x = 10, y = "z"))
```

Use `plyr::rbind.fill()` to combine data frames that don’t have the same columns.

### 1.4.4 Special columns

Since a data frame is a list of vectors, it is possible for a data frame to have a column that is a list:

```{r}
df <- data.frame(x = 1:3)
df$y <- list(1:2, 1:3, 1:4)
df
```

However, when a list is given to data.frame(), it tries to put each item of the list into its own column, so this fails:

```
data.frame(x = 1:3, y = list(1:2, 1:3, 1:4))
> data.frame(x = 1:3, y = list(1:2, 1:3, 1:4))
Error in data.frame(1:2, 1:3, 1:4, check.names = FALSE, stringsAsFactors = TRUE) : 
  arguments imply differing number of rows: 2, 3, 4

```

A workaround is to use I(), which causes data.frame() to treat the list as one unit:

```{r}
dfl <- data.frame(x = 1:3, y = I(list(1:2, 1:3, 1:4)))
str(dfl)
```
Similarly, it’s also possible to have a column of a data frame that’s a matrix or array, as long as the number of rows matches the data frame:
    
```{r}
dfm <- data.frame(x = 1:3, y = I(matrix(1:9, nrow = 3)))
str(dfm)
dfm[2, "y"]
```

### 1.4.5 Exercises

1. What attributes does a data frame possess?

    a data frame has 'names', 'row.names' and 'class' attributes.
    
2. What does as.matrix() do when applied to a data frame with columns of different types?

    Coerce the columns to the same type.

3. Can you have a data frame with 0 rows? What about 0 columns?

    We can have a data frame with 0 rows and 0 columns.
