---
title: "Data Structures"
output: pdf_document
---

## Introduction

R's base data structures can be organized by their dimensionality (1d, 2d, or nd) and whether they're homogeneous (all contents must be of the same type) or heterogeneous (the contents can be of different types). This gives rise the five data types most often used in data analysis:

|    | Homogeneous     |  Heterogeneous  |
|----|:---------------:|----------------:|
| 1d | Atomic vector   | List            |
| 2d | Matrix          | Data frame      |
| nd | Array

Almost all other objects are built upon these foundations. Note that R has no 0-dimentional, or scalar types. Individual numbers or strings are actually vectors of length one. Given an object, the best way to understand what data structures it's composed of is to use `str()`.

## 1.1 vectors
The basic data structure in R is the vector. There are two kinds of vectors: atomic vectors and lists. They have three common properties:   
* Type, typeof(), what it is.
* Length, length(), how many elements it contains.
* Attributes, attributes(), additional arbitrary metadata.

They differ in the types of their elements: all elements of an atomic vector must be the same type, whereas the elements of a list can have different types. Use `is.vector() || is.list()` to test if an object actually a vector.

## 1.1.1 Atomic Vector
There are four common types of atomic vectors: logical, integer, double(often called numeric), and character.

Create atomic vectors with `c()`:
```{r}
dbl_var <- c(1, 2.5, 4.5)
int_var <- c(1L, 6L, 10L)# get an integer vector with L suffix
log_var <- c(TRUE, FALSE, T, F) # use TRUE, FALSE(or T, F) to create logical vector
chr_var <- c("these are", "some strings")

##Atomic vectors are always flat
c(1, c(2, c(3, 4)))
```

### 1.1.1.1 Types and tests
We can determine a vector's type with `typeof()`, or check if it is a specific type with an "is" function: `is.character()`, `is.double()`, `is.integer()`, `is.logical()`, or `is.atomic()`.

```{r}
typeof(int_var)
is.integer(int_var)
is.atomic(int_var)

typeof(dbl_var)
is.double(dbl_var)
is.atomic(dbl_var)
```

`is.numeric` is a general test and return *TRUE* for both integer and double vectors.

### 1.1.1.2 Coercion
If we combine two types of vectors they will be *coerced* to the most flexible type. Types from least to most flexible are: logical, integer, double and character.







